Follow-up Samples

### 2017-02-22 submission/manuscript_outline_20161024.Rmd
incorporate Amanda suggestions results and discussion

***Bacterial Community and FIT:***
We first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma or carcinoma.  What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1a]. We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1b].  The broad shift in bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1d] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].  We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].

***Carcinoma Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  There was a decrease in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A]. In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2B]. We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of any of these respective OTUs [Figure 2].

***The Lesion Model:***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal control individuals.  We investigated this by creating a model to classify lesion versus normal based on the bacterial community and FIT measurements.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].  There were a total of `r length(lesion_top10_varsMDA$variable)` variables that were used in this model [Figure 3B].  The FIT measurement for fecal blood resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].

@@

***The Initial Sample Model:***
After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].  There were a total of `r length(IF_top10_varsMDA$variable)` variables that were used for this model [Figure 4B].  The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].

@@

***Treatment Differences:***
After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].

@@

From our results there were large observed differences in the bacterial microbiome between pre and post lesion removal samples based on whether the individual had an adenoma or carcinoma.  There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure S1].

@@

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).  What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 3 & 4] but also that the commonly associated CRC bacteria were not highly represented within this model.  Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  We followed this up by creating a second model to classify initial versus follow up samples.  We found that this model was able to accurately classify initial versus follow up samples [Figure 3 & 4].  Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S2].  Although there was a detectable change in the carcinoma group and this change was towards what would be expected for normal controls, the follow up samples may not be a completely normal bacterial microbiome.

For the majority of tests performed there were no differences in the bacterial microbiome based on whether chemotherapy or radiation was received [Table S3].  Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).  These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.  Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].

@@

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our lesion model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.  This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.  Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.  The time difference in collection of sample between adenoma and carcinoma could have affected our observations for differences between individuals with adenoma or carcinoma.  This confounding though would not affect the observations where these individuals are grouped together.

Interestingly, within figure 4 the initial sample model showed better test AUC results then the training set AUC.  This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies and similar conclusions can be drawn solely from this model alone.  Further independent studies need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogenesis of disease.

>>>>

***Bacterial Community and FIT:***
Within our `r length(rownames(good_metaf))` person cohort we first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`).  We found that the bacterial community in those with carcinoma were more dissimilar to their initial sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1a]. We also found that there were larger changes in fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value `r ifelse(change_summary['fit', 'Pvalue'] < 0.0001, paste("<"), paste(">"))` 0.0001) [Figure 1b].  The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure1c] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1d] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  When investigating alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].  We also observed that there was no difference in the relative abundance of any OTU between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].

***Carcinoma Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were changes in previously well described carcinoma associated OTUs, such as *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  We first observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance for any of these respective OTUs [Figure 2].  Despite this, those with carcinoma had a decrease in relative abundance from initial to follow up for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 2A]. In contrast, there was no difference in relative abundance in any of these OTUs when considering only those with adenoma [Figure 2B].

***The Lesion Model:***
We next wanted to identify if there were any common bacterial microbiome changes in individuals with adenoma or carcinoma versus normal controls.  We investigated this by creating a model to classify samples as lesion versus normal based on the bacterial community and FIT measurements.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100 different test set AUC iterations [Figure 3A].  There were a total of `r length(lesion_top10_varsMDA$variable)` variables that were used in this model [Figure 3B].  The FIT measurement for fecal blood resulted in the largest decrease in MDA while the OTU with the largest MDA was *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].

@@

***The Initial Sample Model:***
After building a model to classify based on lesion we built a separate model specifically to be able to classify whether samples were initial samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 4A].  There were a total of `r length(IF_top10_varsMDA$variable)` variables that were used for this model [Figure 4B].  The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) while FIT measurement for fecal blood resulted in the sixth largest decrease in MDA [Figure 4B].

@@

***Treatment Differences:***
After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  This suggests that those treated for chemotherapy my have had a larger change from the initial sample then those without such treatment.  All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].

@@

From our results there were large observed differences in the bacterial microbiome in samples before and after lesion removal based on whether the individual had an adenoma or carcinoma.  In individuals with carcinoma compared to adenoma there were much larger differences between initial and follow up samples based on the community beta diversity and in fecal blood as measured by FIT [Figure 1].  However, there were no differences between initial and follow up samples for any alpha diversity metric measured regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure S1].

@@

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).  Based on this model we found that the follow up samples were closer to normal then the initial samples due to a decrease in positive probability for lesion and that the commonly associated CRC bacteria were not highly represented within this model with the exception of *Porphyromonas asaccharolytica* [Figure 3].  Although there was a detectable change and this change was towards what would be expected for normal controls, it should be noted that the follow up samples may not be a completely normal bacterial microbiome.

After creating the lesion mode we then created a second model to classify initial versus follow up samples.  We found that this model was able to accurately classify initial versus follow up samples suggesting that regardless of adenoma or carcinoma there are distinct common changes within the bacterial microbiome that occurs after lesion removal [Figure 4].  Both models had OTUs that overwhelmingly belonged to commensal bacteria.  Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S2].

Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) than carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).  These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.  Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d]. This confounding though would not affect the observations where these individuals were grouped together.

@@

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our lesion model keeping a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.  This study also drew heavily from those with Caucasian ancestry making it possible that the observations may not be representative of those with either Asian or African ancestry.  Although our training and test set are relatively large we still run the risk of over-fitting or having a model that may not be representative of other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

Interestingly, within the initial sample model the test data performed better than the training data.  This may have occurred because the training AUC determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be over-fit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepancies.  Further independent studies need to be carried out to verify our findings and models since we are dealing with correlations that may not be truly representative of the pathogenesis of disease.
